---
title: "Vignette for mlmm package"
author: "Irene SL Zeng"
date: "15 March 2018"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette for mlmm package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Read data
```{r read data,eval=TRUE}
data(pdata,package="mlmm")
```

##Check if there is overlapped definition for missing and censor
```{r table for miss and censor,eval=TRUE}
table(pdata$miss,pdata$censor)
```

##If there are overlapping, convert missing from 1 to 0 where censor =1
```{r re-assign miss and censor,eval=TRUE}
n=dim(pdata)[1]
for (i in seq_len(n)) if (pdata$miss[i]==1 && pdata$censor[i]==1) pdata$miss[i]=0
```

##Set the formula for response, missingness and subject
```{r set formular examples,eval=TRUE}
formula_completed=var1~var2+treatment;
formula_missing=miss~var2;
formula_censor=censor~1;
formula_subject=~treatment;
response_censorlim=0.002;
```

##Set the path and fit the regression model using mlmc()
The formula for subject will need to have at least one variable. For example, to include subject-id for subject and this will provide the mean value for each subject. 
```{r example for using function mlmc() and mlmm(),eval=FALSE}
model1=mlmc(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_censor=censor~1,formula_subject=~sid,pdata=pdata,response_censorlim=0.002,
respond_dep_missing=TRUE,pidname="geneid",sidname="sid",
iterno=100,chains=2,savefile=TRUE,usefit=FALSE)

model2=mlmc(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_censor=censor~1,formula_subject=~sid,pdata=pdata,response_censorlim=0.002,
respond_dep_missing=TRUE,pidname="geneid",sidname="sid",
iterno=1000,chains=2,savefile=TRUE,usefit=TRUE)

model3=mlmc(formula_completed=var1~var2,formula_missing=miss~var2,
formula_censor=censor~1,formula_subject=~sid+treatment,pdata=pdata,response_censorlim=0.002,
respond_dep_missing=TRUE,pidname="geneid",sidname="sid",
iterno=100,chains=2,savefile=TRUE,usefit=FALSE)

model3b=mlmc(formula_completed=var1~var2,formula_missing=miss~var2,
formula_censor=censor~1,formula_subject=~sid+treatment,pdata=pdata,response_censorlim=0.002,
respond_dep_missing=TRUE,pidname="geneid",sidname="sid",
iterno=1000,chains=2,savefile=TRUE,usefit=TRUE)

##using mlmm()function. Formular for completed data, subject and missingness requires at least one variable
##r example for using function mlmm()
model4=mlmm(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_subject=~sid,pdata=pdata,respond_dep_missing=FALSE,pidname="geneid",sidname="sid",
iterno=100,chains=2,savefile=TRUE,usefit=FALSE)

model5=mlmm(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_subject=~sid,pdata=pdata,respond_dep_missing=FALSE,pidname="geneid",sidname="sid",
iterno=1000,chains=2,savefile=TRUE,usefit=TRUE)

model5=mlmm(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_subject=~sid+treatment,pdata=pdata,respond_dep_missing=FALSE,pidname="geneid",sidname="sid",
iterno=100,chains=2,savefile=TRUE,usefit=FALSE)

model5b=mlmm(formula_completed=var1~var2+treatment,formula_missing=miss~var2,
formula_subject=~sid+treatment,pdata=pdata,respond_dep_missing=FALSE,pidname="geneid",sidname="sid",
iterno=1000,chains=2,savefile=TRUE,usefit=TRUE)

```

## Plot the parameters
## From the print out, we can see that, if the estimate converge, it will have high number of efficient samples and the r value close to 1 
The following codes are to 
1) plot posterior parameter using the outsummary results from multiple chains and, 
2) plot the trajectory including those values generated from burn-in iterations.

```{r plot one,echo=TRUE,eval=FALSE}
summaryreader=read.csv(file=file.path(getwd(),"outsummary.csv"),header=T,sep=",",skip=0)
iterno=dim(summaryreader)[1]
burnin=iterno/2
U.1.1=rowMeans(matrix(c(summaryreader$chain.1.U.1.1,summaryreader$chain.2.U.1.1),nrow=iterno,ncol=2))[burnin:iterno]
meanU=mean(U.1.1)
qU=quantile(U.1.1,p=seq(0,1,by=0.025))
scale=seq(0,1,by=0.025)
plot(scale,qU,pch=19,ylab="quantiles of estimate",xlab="quantiles")

segments(0,qU[names(qU)=="50%"],1,qU[names(qU)=="50%"],lwd=2,col="red")
segments(0,qU[names(qU)=="2.5%"],1,qU[names(qU)=="2.5%"],lty=2,lwd=2,col="red")
segments(0,qU[names(qU)=="97.5%"],1,qU[names(qU)=="97.5%"],lty=2,lwd=2,col="red")
legend(0.5,qU[names(qU)=="50%"],"median",cex=0.8,bty="n")
legend(0.03,qU[names(qU)=="2.5%"],"2.5%",cex=0.8,bty="n")
legend(0.90,qU[names(qU)=="97.5%"],"97.5%",cex=0.8,bty="n")
qU
```

```{r plot two,echo=TRUE,eval=FALSE}
sample1reader=read.csv(file=file.path(getwd(),"samples_1.csv"),header=T,sep=",",skip=25)
sample2reader=read.csv(file=file.path(getwd(),"samples_2.csv"),header=T,sep=",",skip=25)

#plot variable U.1.1 - the intercept of first unit
trajectory_length=dim(sample1reader)[1]
plot(seq(1,trajectory_length,by=1),sample1reader$U.1.1,xlab="trajectory number",ylab="U.1.1",type="n",ylim=c(min(sample1reader$U.1.1,sample2reader$U.1.1,na.rm=TRUE),max(sample1reader$U.1.1,sample2reader$U.1.1,na.rm=TRUE)))
trajectory=seq(1,trajectory_length,by=1)
lines(trajectory,sample1reader$U.1.1)
lines(trajectory,sample2reader$U.1.1,col="red")
```


